---
title: "Guidance on the Madhava Graph Problem"
author: "Homer White"
date: 2017-09-27T10:30:00
tags: ["csc115", "flow control"]
aggregators: [""]
---

```{r include=FALSE}
knitr::opts_chunk$set(tidy=FALSE,out.width="90%", fig.align='center')
library(ggplot2)
```

## Introduction

This post provides a bit of guidance for students in CSC 115 who are working on the [Madhava Graph exercise](http://homerhanumat.com/r-notes/exercises-2.html) in Chapter 4 of the course text.

<!--more-->

As you will recall, the statement of the problem is as follows:

>Recall the function `madhavaPI()` from [Section 3.4.1](). Use this function to write a function called `madhavaGraph()` that will do the following: given a number $n$, the function uses **ggplot2** to produce a line graph of the first $n$ approximations to $\pi$ using the initial terms of the Madhava series.

>The plot should be a line graph similar to the one produced by the `collatz()` functions from this Chapter. The function should take a single argument `n`, whose default value is 10.

>It should validate the input: if the number entered is not at least 1, then the function should should explain to the user that the he/she must enter a positive number, and then stop.


## Outline

Based on the specification in the problem, you can set up an initial outline of the desired function, as follows:

```{r eval=FALSE}
madhavaGraph <- function(n = 10) {
  
  validate the input: if n isn't at least 1, stop the function
    
  find the sums:
    * the sum of the first term (4)
    * the sum of the first and second terms (4 - 4/3)
    * the sum of the first three terms ( 4 - 4/3 + 4/5)
    * and so on until ...
    * the sum of the first n terms
  make sure these sums are stored in a vector (let's call it "results")
  
  make the plot:
    - the vector 1:n (call it "terms") gives the x-coordinates of the points
    - the results vector gives y-coordinates
  
}
```

As you can see, the body of the function has three primary parts:

1. validating the input
1. finding the sums and storing them in the `results` vector
1. making the plot.

I'll leave Part 1 entirely to you, but say a bit more about parts 2 and 3.

## Making the Graph

First we'll think about the graph.  Since it is made from functions in the **ggplot2** package, you will need to make sure that you load the package when you are designing the function.`r tufte::margin_note("The function itself does not load **ggplot2**, so you should make sure to tell anyone who plans to use the function to load it!")`

```{r eval=FALSE}
library(ggplot2)
```

Now suppose that the value of the parameter `n` is 10.  Then by the time you get around to making the graph you'll have the terms:

```{r echo=FALSE}
n <- 10
```


```{r}
terms <- 1:n
terms
```

You'll also have the sums, another vector of length 10:

```{r echo=FALSE}
madhavaPI <- function(n = 1000000) {
  k <- 1:n
  terms <- (-1)^(k+1)*4/(2*k-1)
  sum(terms)
}
n <- 10
results <- numeric(n)
for ( i in 1:n) {
  results[i] <- madhavaPI(n = i)
}
```


```{r}
results
```

From our study of the Collatz function, we know that the way to make a line graph of these points is:

```{r}
plotTitle <- paste0("The first ", n, " Madhava sums")
  p <- ggplot(mapping = aes(x = terms, y = results)) +
    geom_point() + geom_line() +
    labs( x = "number of terms", y = "Madhava sum",
          title = plotTitle)
  print(p)
```

Hence we can fill in the outline of our function a bit:

```{r eval=FALSE}
madhavaGraph <- function(n = 10) {
  
  validate the input: if n isn't at least 1, stop the function
    
  find the sums:
    * the sum of the first term (4)
    * the sum of the first and second terms (4 - 4/3)
    * the sum of the first three terms ( 4 - 4/3 + 4/5)
    * and so on until ...
    * the sum of the first n terms
  make sure these sums are stored in a vector (let's call it "results")
  
 plotTitle <- paste0("The first ", n, " Madhava sums")
  p <- ggplot(mapping = aes(x = terms, y = results)) +
    geom_point() + geom_line() +
    labs( x = "number of terms", y = "Madhava sum",
          title = plotTitle)
  print(p)
  
}
```


## Finding and Storing the Sums

The outline for this part of the task looked  like:

* get the sum of the first term (just 1)
* get the sum of the first and second terms ($4 - 4/3$)
* get the sum of the first three terms ($4 - 4/3 + 4/5$)
* and so on until ...
* get the sum of the first n terms

At each step in above process, we will need to store our sum in the `results` vector, so that it isn't lost.

The natural way to accomplish the above task is to use a `for`-loop in which we iterate by index and store results.  The outline of such a loop is:

```{r eval=FALSE}
results <- numeric(n)
for ( i in 1:n ) {
  
  results[i] <- the sum of the first i terms
  
}
```

We already know how to find the sum of the first $i$ terms:  we can just use the function `madhavaPI()` from Chapter 3:

```{r}
madhavaPI <- function(n = 1000000) {
  k <- 1:n
  terms <- (-1)^(k+1)*4/(2*k-1)
  sum(terms)
}
```

Hence when you are working inside the loop, you can get and store the results all in one line:

```{r eval=FALSE}
results[i] <- madhavaPI(i)
```

Well, to be totally truthful, you can do this *provided that* R knows what `madhavaPI()` is.  Hence you must either:

* hope that whoever uses `madhavaGraph()` has `madhavaPI()` in their Global Environment so that `madhavaGraph()` can look up the name `madhavaPI` when it needs it, or
* define `madhavaPI()` inside the body of `madhavaGraph()`, before you start computing any sums.

The latter seems to be the safer choice!

## Put It All Together

You still have the task of filling in the outline completely and then testing `madhavaGraph()` until it works.





