---
title: "Exploding Boxplots in R"
author: "Homer White"
date: 2018-01-31T10:30:00
tags: ["htmlwidgets", "javascript"]
aggregators: ["r-bloggers"]
excerpt: "The package bpexploder provides an HTML Widget that makes boxplots that explode upon mouse-click into individual-value plots.  The plots allow for a modest amount of customization, including tool-tips."
ghcommentid: 4
---

```{r include=FALSE}
library(bpexploder)
```

<p id="reference"></p>

`bpexploder` `r tufte::margin_note("**Credits**:  The JavaScript library is based on Mathieu Caule's [D3 Exploding Boxplots](https://mcaule.github.io/d3_exploding_boxplot/), which I have modified slightly and updated for D3 Version 4. The tool-tips were created by [Justin Palmer](https://github.com/Caged), updated to D3v4 by [Dave Gotz](https://github.com/VACLab/d3-tip) and tweaked slightly by myself.")`represents the author's initial foray into Html Widgets for R; it renders an svg element showing box-plots that explode upon mouse-click into jittered individual-value plots.  The user has the option to configure tool-tips for the individual points.

## Installation and Usage

Install the package from Git Hub:

```{r eval = FALSE}
devtools::install_github("homerhanumat/bpexploder")
```



Call the function with the notorious `iris` data:

```{r eval = FALSE}
bpexploder(data = iris,
           settings = list(
             groupVar = "Species",
             levels = levels(iris$Species),
             yVar = "Petal.Length",
             tipText = list(
               Petal.Length = "Petal Length"
             ),
             relativeWidth = 0.75)
           )
```

```{r echo = FALSE}
bpexploder(data = iris,
           settings = list(
             groupVar = "Species",
             levels = levels(iris$Species),
             yVar = "Petal.Length",
             tipText = list(
               Petal.Length = "Petal Length"
             ),
             relativeWidth = 0.75,
             referenceId = "reference",
             align = "left")
           )
```

Click to explode boxes, double-click to restore.

##  Settings

`bpexploder` provides modest options for customization, as illustrated in the following example:

```{r eval = FALSE}
bpexploder(data = chickwts,
           settings = list(
             yVar = "weight",
             # default NULL would make make one plot for yVar
             groupVar = "feed",
             levels = levels(with(chickwts,
                                  reorder(feed, weight, median))),
             # you could adjust the group labels ...
             levelLabels = NULL,
             # ... and the colors for each group:
             levelColors = RColorBrewer::brewer.pal(6, "Set3"),
             yAxisLabel = "6-week weight (grams)",
             xAxisLabel = "type of feed",
             tipText = list(
               # as many os you like of:
               # variableName = "desired tool-tip label"
               # leave tipText at NULL for no tips
               weight = "weight"),
             # set width relative to grandarent element of svg image:
             relativeWidth = 0.75,
             # default alignment within containing div is "center"
             # "left" and "right" are other possible values"
             align = "left",
             # aspect = width/height, defaults to 1.25
             aspect = 1.5)
           )
```

```{r echo = FALSE}
bpexploder(data = chickwts,
           settings = list(
             yVar = "weight",
             # default NULL would make make one plot for yVar
             groupVar = "feed",
             levels = levels(with(chickwts,
                                  reorder(feed, weight, median))),
             # you could adjust the group labels ...
             levelLabels = NULL,
             # ... and the colors for each group:
             levelColors = RColorBrewer::brewer.pal(6, "Set3"),
             yAxisLabel = "6-week weight (grams)",
             xAxisLabel = "type of feed",
             tipText = list(
               # as many os you like of:
               # variableName = "desired tool-tip label"
               # leave tipText at NULL for no tips
               weight = "weight"),
             # set width relative to grandarent element of svg image:
             relativeWidth = 0.75,
             # default alignment within containing div is "center"
             # "left" and "right" are other possible values"
             align = "left",
             # aspect = width/height, defaults to 1.25
             aspect = 1.5,
             referenceId = "reference")
           )
```


## Sizing

By default the box-plot chart sizes itself so that its width is the the setting `relativeWidth` (default-value 1) multiplied by the offset-width of its grandparent node in the HTML DOM.  For some layouts---including the `tuftesque` theme of this post---the default might not give you the type of control you want.  In that case you may direct `epexploder()` to make the width of the chart track the offset-width of an existing DOM element.  For example, if paragraphs in your document have the desired width, then create an empty paragraph in your markdown like this:

```
<p id="refelem"></p>
```

To set the widget-width to a specific fraction of the element that contains the paragraph, you could again use the `relativeWidth` setting, or you might try something like: 

```
<p id="refelem" style="width: 70%"></p>
```

<p id="refelem"></p>

In either case, call `bpexploder()` with the `referenceId` setting: `r tufte::margin_note("In fact I've been using a reference paragraph throughout this post.  Note also that *align = \"left\"* is important in **tuftesque** and probably in other Tufte-based themes.")`

```{r}
bpexploder(data = iris,
           settings = list(
             groupVar = "Species",
             levels = levels(iris$Species),
             yVar = "Petal.Length",
             tipText = list(
               Petal.Length = "Petal Length"
             ),
             #using <p id="refelem"></p>
             referenceId = "refelem",
             relativeWidth = 0.75,
             align = "left")
           )
```

## Known Issues and Limitations

* Using the Leonids theme from the `prettydoc` R Markdown package, tool-tips are not visible in Firefox or Safari.  (They do show in Chrome.)
* When `htmlwidgets::createWidget()` creates a widget it gives the widget a random Id number.  For some reason this can result in spurious warnings concerning the hidden variable `.Random.Seed`.  To work around this, set a seed in the `setup` chunk of your R Markdown document, e.g.:

```{r eval = FALSE}
library(bpexploder)
set.seed(5437)
```